\Large Carryover effects in free recall reveal how past experiences influence memories of future experiences

### 2023-01-17 paper/main.tex
added a discussion section on expectation, event boundaries, and situation models

A key mystery is whether (and how) the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal22, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example by sorting the words in a random order~\citep{Kaha12}.

@@

>>>>

A key mystery is whether (and how) the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal23, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example by sorting the words in a random order~\citep{Kaha12}.

@@

Our findings that participants' current and future memory behaviors are
sensitive to manipulations in which features change over time, and how features
change across items and lists, suggest parallels with studies on how we form
expectations and predictions, segment our continuous experiences into discrete
events, and make sense of different scenarios and situations. Each of these
real-world cognitive phenomena entail identifying statistical regularities in
our experiences, and exploiting those regularities to gain insight, form
inferences, organize or interpret memories, and so on. Our past experiences
enable us to predict what is likely to happen in the future, given what
happened ``next'' in our previous experiences that were similar to
now~\citep{Mann20, EichFort09, BarrEtal20, Brig12, ChowEtal16, GlucEtal02,
GoldEtal21, GrifStey03, JonePash07, KimEtal14, TamiThor18, XuEtal23}.

When our expectations are violated, such as when our observations disagree with
our predictions, we may perceive the ``rules'' or ``situation'' to have
changed. \textit{Event boundaries} denote abrupts change in the state of our
experience, for example when we transition from one situation to
another~\citep{RadvZack17, ZwaaRadv98}. Crossing an event boundary can impair
our memory for pre-boundary information and enhance our memory for
post-boundary information~\citep{RadvCope06, SahaKell02, MannEtal16}. Event
boundaries are also tightly associated with the notion of \textit{situation
models} and \textit{schemas}-- mental frameworks for organizing our
understanding about the rules of how we and others are likely to behave, how
events are likely to unfold over time, how different elements are likely to
interact, and so on. For example, a situation model pertaining to a particular
restaurant might set our expectations about what we are likely to experience
when we visit that restaurant (e.g., what the building will look like, how it
will smell when we enter, how crowded the restaurant is likely to be, the
sounds we are likely to hear, etc.). Similarly, we might learn a schema
describing how events are likely to unfold \textit{across} any sit-down
restaurant-- e.g., open the door, wait to be seated, receive a menu, decide
what to order, place the order, and so on. Situation models and schemas can
help us to generalize across our experiences, and to generate expectations
about how new experiences are likely to unfold. When those expectations are
violated, we can perceive ourselves to have crossed into a new situation.

In our study, we found that abruptly changing the ``rules'' about how the
visual appearances of words are determined, or about the orders in which words
are studied, can lead participants to behave similarly to what one might expect
upon crossing an event boundary. Adding in variability in font color and
presentation locations for words on late lists, after those visual features had
been held constant on early lists, led participants to remember more words on
those later lists. (We found the same pattern when we varied words' colors and
locations on early lists but held them constant on late lists.) One potential
explanation is that participants perceive an ``event boundary'' to have
occurred when they encounter the first ``late'' list. According to contextual
change accounts of memory across event boundaries~\citep[e.g.,][]{SahaKell02},
this could help to explain why participants in the reduced (early) and reduced
(late) conditions exhibited better overall memory performance. Specifically,
their memory for late list items could benefit from less interference from
early list items, and the contextual features associated with late list items
(after the ``event boundary'') might serve as more specific recall cues for
those late items (relative to if the boundary had not occurred).
